# How Chemical Properties Of Red Wine Influence Quality
### by Kyle Campbell

-------------------------------------------------------------------------

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(tidyr)
library(dplyr)
library(GGally)
library(knitr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
redwine <- read.csv('red wine.csv')
```


## Introduction to the Data Set 

The red wine dataset that I chose contains 1,599 red wines with 12 variable
related to the chemical properties of each wine. Three wine experts were 
responsible for rating the quality of each wine with a scale from 0 (terrible)
up to 10 (exceptional).

## Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
summary(redwine)
```

```{r echo=FALSE}
names(redwine)
```

After taking an initial look at our data, I want to explore some of the counts
of these variables such as alcohol, quality, and pH. I would like to look at
the distributions to identify and possibly remove outliers from some of the 
variables so that we have a more accurate analysis.

## Distribution of All Variables

```{r echo=FALSE}
v1 <- ggplot(data = redwine, aes(x = fixed.acidity))+
        xlab('Fixed Acidity')+
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.5)

v2 <- ggplot(data = redwine, aes(x = volatile.acidity)) +
        xlab('Volatile Acidity') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.05)

v3 <- ggplot(data = redwine, aes(x = citric.acid)) +
        xlab('Citric Acid') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.05)

v4 <- ggplot(data = redwine, aes(x = residual.sugar)) +
        xlab('Residual Sugar') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.4)

v5 <- ggplot(data = redwine, aes(x = chlorides)) +
        xlab('Chlorides') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.025)

v6 <- ggplot(data = redwine, aes(x = free.sulfur.dioxide)) +
        xlab('Free Sulfur Dioxide') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 2)

v7 <- ggplot(data = redwine, aes(x = total.sulfur.dioxide)) +
        xlab('Total Sulfur Dioxide') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 10)

v8 <- ggplot(data = redwine, aes(x = density)) +
        xlab('Density') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.001)

v9 <- ggplot(data = redwine, aes(x = pH)) +
        xlab('pH') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.05)

v10 <- ggplot(data = redwine, aes(x = sulphates)) +
        xlab('Sulphates') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.1)

v11 <- ggplot(data = redwine, aes(x = alcohol)) +
        xlab('Alcohol') +
        geom_histogram(colour='black', fill='green', 
                       binwidth = 0.2)


grid.arrange(v1, v2, v3, v4, v5, v6, v7, v8, v9, v10, v11, ncol=3)
```

I am surprised to find that a larger portion of the wines sampled tend to skew
towards under 10% alcohol by volume. This visual confirms the guidelines of 
wine, which state they must be between 9-14%. 

The variables which seem to have a normal distribution are **Volatile Acidity**,
**pH**, and **Density**. The variables with a positively skewed distribution are
**Fixed Acidity**, **Citric Acid**, **Free Sulfur Dioxide**, **Total Sulfur** 
**Dioxide**, and **Alcohol**. The variables that seem to have a long tail 
distribution are **Residual Sugar**, **Sulphates**, and **Chlorides**.

I would like to further inspect the three variables with long tail 
distributions to identify outliers.

```{r echo=FALSE}

b1 <- ggplot(data = redwine, aes(x = 1, y = residual.sugar)) + 
              geom_jitter(alpha = 0.1 ) +
              geom_boxplot(alpha = 0.2, color = 'red') 

b2 <- ggplot(redwine, aes( x   = residual.sugar)) + 
              geom_histogram(bins=30 )

b3 <- ggplot(data = redwine, aes(x = 1, y = sulphates)) + 
              geom_jitter(alpha = 0.1 ) +
              geom_boxplot(alpha = 0.2, color = 'red') 

b4 <- ggplot(redwine, aes( x   = sulphates)) + 
              geom_histogram(bins=30 )

b5 <- ggplot(data = redwine, aes(x = 1, y = chlorides)) + 
              geom_jitter(alpha = 0.1 ) +
              geom_boxplot(alpha = 0.2, color = 'red') 

b6 <- ggplot(redwine, aes( x   = chlorides)) + 
              geom_histogram(bins=30 )

grid.arrange(b1, b2, b3, b4, b5, b6, ncol = 2)
```

Here we can see the outliers at the upper end of each of these three variables. 
I am going to leave these outliers in the dataset, however it's good to be 
aware that they are present when examining the plots utilizing these variables.

For the rest of the variables containing positively skewed distributions, I will
try to do a log transformation on them to attempt to normalize their 
distribution.

```{r echo=FALSE, warning=FALSE}

redwine$log.fixed.acidity <- log(redwine$fixed.acidity)
l1 <- ggplot(data = redwine, aes(x = log.fixed.acidity))+
        xlab('Fixed Acidity(log)')+
        geom_histogram(colour='black', fill='blue', 
                       binwidth = 0.05)

redwine$log.citric.acid <- log(redwine$citric.acid)
l2 <- ggplot(data = redwine, aes(x = log.citric.acid)) +
        xlab('Citric Acid(log)') +
        geom_histogram(colour='black', fill='blue', 
                       binwidth = 0.1)

redwine$log.free.sulfur.dioxide <- log(redwine$free.sulfur.dioxide)
l3 <- ggplot(data = redwine, aes(x = log.free.sulfur.dioxide)) +
        xlab('Free Sulfur Dioxide(log)') +
        geom_histogram(colour='black', fill='blue', 
                       binwidth = 0.3)

redwine$log.total.sulfur.dioxide <- log(redwine$total.sulfur.dioxide)
l4 <- ggplot(data = redwine, aes(x = log.total.sulfur.dioxide)) +
        xlab('Total Sulfur Dioxide(log)') +
        geom_histogram(colour='black', fill='blue', 
                       binwidth = 0.3)

grid.arrange(l1, l2, l3, l4, ncol=2)
```

The distributions look a little more normalized with a log transformation 
applied. We will use these transformed variables in our bivariate and 
multivariate analysis. The citric acid variable has over 100 values showing "0".
I'm not sure if this is due to values not being available, or if those values
are truly zero. I'm going to remove the log transformed citric acid variable and
also remove the original variables columns for the characteristics with
a successful log transformation.

```{r echo = TRUE}
redwine$log.citric.acid <- NULL
redwine$fixed.acidity <- NULL
redwine$free.sulfur.dioxide <- NULL
redwine$total.sulfur.dioxide <- NULL
```

Let's finally get a look at the quality variable.

```{r echo=FALSE}
qplot(data = redwine, x = quality, binwidth = 0.5)
```

It appears that even though our reviews were able to scale from 1-10 based on 
quality, they kept the quality ratings between 3-8. It makes sense that the 
majority of red wine samples fall in the 5-6/average rating. I am going to 
create a new variable to put the ratings in buckets with 0-4 being "bad", 
5-6 being "average", and 7-8 being "good".

```{r echo=TRUE}
redwine$rating <- ifelse(redwine$quality < 5, "bad",
                         ifelse(redwine$quality > 6, "good", "average"))

redwine$rating <- ordered(redwine$rating,
                     levels = c('bad', 'average', 'good'))

qplot(data = redwine, x = rating)
```

Here we can see that the majority, **over 1250**, of red wine samples were
rated as average. Having such a large portion of our dataset rated as
average red wine samples may make it harder for us to identify correlations
between what qualities make a good wine, but we will try to focus on the 
good and bad rating.

```{r echo=FALSE}
qplot(data = redwine, x = density, binwidth = 0.0001)
```

I wanted to get a closer look at the plot for the density variable. I'm curious 
to see how density and alchol are correlated. We will explore this further
in the bivariate analysis.

## Univariate Analysis


### What is the structure of your dataset?
This dataset contains 1599 observations of red wine samples with 13 variables
for each observation. All variables consist of numerical data, except for our 
quality variable which is categorical. I transformed the quality variable
to only consist of three ratings, "bad", "average", and "good" to more
easily identify patterns correlating with certain variables.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in my dataset is the quality of the red wine
sample.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest
I think the alcohol percentage, as well as the chemical composition of the 
alcohol, will both play a huge role in supporting the investigation into my 
feature of interest.

### Did you create any new variables from existing variables in the dataset?
I created a new variable from the quality data to give each wine a rating. 
Since most of the data was clustered around 5-6, I gave that an "average" 
rating. 0-4 I gave a "bad" rating and 7-10 I gave a "good" rating.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The red wine dataset was mostly tidy. I added an additional column to put the
quality variable into buckets categorized by rating. I also transformed a few 
of the variables using a log transformation to try to normalize the 
distribution, as well as pointing out a few outliers with some of the 
variables in the red wine dataset.


## Bivariate Plots Section

Let's first check the correlations of every variable matched up
against each other to try to find out which ones have high/low correlations to 
each other.

```{r echo=FALSE, Plot_Three, warning=FALSE}
ggcorr(redwine, label = TRUE) +
         ggtitle('Correlation of All Variables') +
         theme(plot.title = element_text(hjust = 0.5))
```


With this correlation matrix, we get a complete overview of variables and their
positive or negative correlations. Right now I would like to focus on the 
variables that are correlated to quality.

#### Top 4 Correlations to Red Wine Quality

Chemical Property | Correlation (Pearson)
-------------     | -------------
Alcohol           |  0.48
Volatile Acidity  | -0.39
Sulphates         |  0.25
Citric Acid       |  0.23

As expected, alcohol has the highest correlation with quality at 0.48. Volatile
acidity has an inverse correlation with quality, and sulphates and citric acid 
have lower correlations with quality yet still a noticable amount. After a quick
google search, it appears that volatile acidity has a huge impact on aroma and 
taste, explaining why it trails right behind alcohol regarding correlations to 
red wine quality.

Let's plot alcohol against the rating variable I created to get a better look
at the correlation between the two. I will just be using the bad and good
rating to get a better picture of the correlation.

```{r echo=FALSE}
ggplot(data = subset(redwine, rating == 'bad' | rating == 'good'),
       aes(x = rating, y = alcohol)) +
       geom_jitter(alpha = 0.5) +
       geom_boxplot(alpha = 0.5, color = 'blue') +
       stat_summary(fun.y = "mean", 
                    geom = "point", 
                    color = "red", 
                    shape = 1, 
                    size = 3)
```

Here we show boxplots with the mean plotted for the red wine samples rated 'bad'
and 'good'. These box plots show the stark contrast between average alcohol 
content in a bad tasting red wine and a good tasting red wine.

Let's take a look at 'bad' and 'good' rated red wines against the other top 
three correlated variables.

```{r echo=FALSE}
bp1 <- ggplot(data = subset(redwine, rating == 'bad' | rating == 'good'),
       aes(x = rating, y = volatile.acidity)) +
       geom_jitter(alpha = 0.5) +
       geom_boxplot(alpha = 0.5, color = 'blue') +
       stat_summary(fun.y = "mean", 
                    geom = "point", 
                    color = "red", 
                    shape = 1, 
                    size = 3)

bp2 <- ggplot(data = subset(redwine, rating == 'bad' | rating == 'good'),
       aes(x = rating, y = sulphates)) +
       geom_jitter(alpha = 0.5) +
       geom_boxplot(alpha = 0.5, color = 'blue') +
       stat_summary(fun.y = "mean", 
                    geom = "point", 
                    color = "red", 
                    shape = 1, 
                    size = 3)

bp3 <- ggplot(data = subset(redwine, rating == 'bad' | rating == 'good'),
       aes(x = rating, y = citric.acid)) +
       geom_jitter(alpha = 0.5) +
       geom_boxplot(alpha = 0.5, color = 'blue') +
       stat_summary(fun.y = "mean", 
                    geom = "point", 
                    color = "red", 
                    shape = 1, 
                    size = 3)

grid.arrange(bp1, bp2, bp3, ncol=2)
```

Our first boxplot shows the inverse relationship that volatile acidity has with 
wine quality. Next, we see that sulphates and citric acid have a positive
correlation with our 'good' and 'bad' wine rating.

Now let's take a quick look at the relationship between alcohol and density
that we proposed earlier.

```{r echo=FALSE}
ggplot(data = redwine, aes(x = density, y = alcohol)) +
  geom_point(alpha=0.2) +
  geom_smooth(method = "lm", se = FALSE, size=1)
```

Here we can see the inverse correlation between alcohol and density. It almost
seems intuitive that as the alcohol % decreases, the density increases. This
is most likely due to alcohol having less density than water, or any other 
chemical components that make up our red wine dataset.

## Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
It was interesting, yet not surprising, to see that alcohol had the highest 
positive correlation with quality/rating.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
I found an inverse correlation with the quality of the red wine and the 
volatile acidity. This makes sense considering the documentation of the 
dataset noted that too high of a reading for volatile acidity will give the
wine an unpleasant, vinegar taste.

### What was the strongest relationship you found?
The strongest relationship found so far was the relationship of alcohol content
to the quality/rating from the judges.

## Multivariate Plots Section

Since I singled in on comparing the 'bad' and 'good' ratings in the bivariate 
analysis, I now want to include all the data for the red wines in my
multivariate analysis. I will start by viewing the relationship between quality,
volatile acidity, and alcohol content.

```{r echo=FALSE}
redwine$alcohol_bucket = cut(redwine$alcohol,
                            c(8, 9, 10.5, 12, 16))

ggplot(data = redwine, aes(x = factor(quality),
           y = volatile.acidity)) + 
       geom_boxplot( aes(fill= alcohol_bucket))  +
       scale_fill_brewer(type='seq', palette = 3,
                   guide=guide_legend(title='Alcohol Content'))
```

Adding in volatile acidity as a variable with alcohol shows the clear effect 
both have on the quality rating. We can easily see the inverse correlation 
between volatile acidity and quality. It seems that the sweet spot for a good
tasting red wine is an alcohol content higher than 12% and a volatile 
acidity lower than 0.6 g/dm^3 .

I would like to take a look at the same visualization with sulphates instead
of volatile acidity to see if the weaker correlation is still visual.

```{r echo=FALSE}
ggplot(data = redwine, aes(x = factor(quality),
           y = sulphates)) + 
       geom_boxplot( aes(fill= alcohol_bucket))  +
       scale_fill_brewer(type='seq', palette = 4,
                   guide=guide_legend(title='Alcohol Content'))
       
```

We can definitely notice the positive correlation between slightly higher 
sulphate levels and increase in quality. We can also see a lot of outliers
with higher readings of sulphates.

Finally let's view how density and alcohol relate to quality.

```{r echo=FALSE}
ggplot(aes(x = density, y = alcohol, 
           color = factor(quality)), 
       data = redwine) +
       geom_point(alpha = 0.8, size = 1) +
       geom_smooth(method = "lm", se = FALSE, size=1)  +
       scale_color_brewer(type='seq', palette = 5,
                   guide=guide_legend(title='Quality'))
```

Our regression lines show the very strong correlation between density and 
alcohol content. In addition, we also see how they both correlate to the 
overall quality of the red wine. The areas in the upper left portion with a 
lower density and higher alcohol content seem to be optimal quality.

## Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Adding the two highest correlating variables of alcohol content and volatile
acidity further reinforced their relationship to the quality of the red wine.

### Were there any interesting or surprising interactions between features?
It was interesting to find the relationship between alcohol content and density,
as well as the relationship between the three variables with the highest
correlation to quality.

------

## Final Plots and Summary

### Plot One
```{r echo=FALSE}
ggplot(data = redwine, aes(x = rating, y = alcohol)) +
       geom_jitter(alpha = 0.2) +
       geom_boxplot(alpha = 0.5, color = 'blue') +
       stat_summary(fun.y = "mean", 
                    geom = "point", 
                    color = "red", 
                    shape = 3, 
                    size = 3) +
       labs(y = 'Alcohol (%)', x = 'Rating') +
       ggtitle('Impact of Alcohol Content on Rating') +
       theme(plot.title = element_text(hjust = 0.5))
```

### Description One
This first plot introduces the huge correlation between alcohol content and
the 'rating' categorical variable I created. In this final plot I show
all three ratings of 'bad', 'average', and 'good' red wines. We can also see 
how our dataset favored data points that fell into the 'average' rating.

### Plot Two: Relationship of Alcohol and Density with Quality
```{r echo=FALSE}
ggplot(aes(x = density, y = alcohol, 
           color = factor(quality)), 
       data = redwine) +
       geom_point(alpha = 0.3, size = 2) +
       geom_smooth(method = "lm", se = FALSE, size=1)  +
       scale_color_brewer(type='seq', palette = 3,
                   guide=guide_legend(title='Quality')) +
       labs(y = 'Alcohol (%)', x = 'Density (g/cm^3)') +
       ggtitle('Relationship of Alcohol & Density w/ Quality') +
       theme(plot.title = element_text(hjust = 0.5))
```

### Description Two
This second plot exemplifies the high correlation between density and alcohol 
content paired with how they both impact quality. This visualization
shows that the sweet spot to be in is a lower density, which correlates to a 
higher alcohol content.

### Plot Three: Top Two Correlated Variables to Quality
```{r echo=FALSE}
redwine$alcohol_bucket = cut(redwine$alcohol,
                            c(8, 9, 10.5, 12, 16))

ggplot(data = redwine, aes(x = factor(quality),
           y = volatile.acidity)) + 
       geom_boxplot( aes(fill= alcohol_bucket))  +
       scale_fill_brewer(type='seq', palette = 3,
                   guide=guide_legend(title='Alcohol Content (%)')) +
       labs(y = 'Volatile Acidity (acetic acid - g / dm^3)', x = 'Quality') +
       ggtitle('How Alcohol Content and Volatile Acidity Effect Quality') +
       theme(plot.title = element_text(hjust = 0.5))
```

### Description Three

This plot may be the most important in our whole analysis. With this 
visualization, we are able to see how the two most highly correlated variables
of alcohol and volatile acidity impact the quality of the red wine sample.


## Reflection

With this exploratory data analysis on the red wine dataset, I found that the 
biggest challenge was understanding the context of the variables I was 
analyzing. With a little help from Google, I was able to get a better 
understanding of what each variable was, and how they relate to the composition
of each of the red wine samples. 

My initial goal was to understand what variables were most important in 
determining the quality of red wine. The visuals and stats I ran on the dataset
started giving me insight into a number of correlations between variables. The 
largest correlation I found in relation to the quality of red wine was the 
alcohol content. I found that the red wines with a higher rating generally had a
higher alcohol content, and the lower quality wines consisted of a lower alcohol
content. This correlation does not necessarily mean that every red wine with a 
high alcohol content tastes good. I believe it's all about finding a sweet spot,
no pun intended, when fermenting a good red wine. Another surprising correlation
was that of citric acid and volatile acidity, which had a large inverse 
relationship.

In the future, running an analysis with the same variables but for different 
types of wine could be beneficial into gaining insight into exactly what makes
the perfect wine. It could maybe also shed some light that there are certain 
chemical properties and factors which determine a good tasting wine depending 
on what variety of wine you are fermenting and processing. This project gave me
a greater appreciation for always asking questions and exploring the data to try
to find answers.


